---
title: "FHIR for Research - Exercise 1 Appendix: Manually querying FHIR endpoints"
output:
  html_document:
    df_print: paged
---

While `fhircrackr` provides an easy method for ingesting data from a FHIR server into R data frames, the `fhircrackr` library does not support all possible requests that can be made to a FHIR server. If you need to make a request that isn't supported by `fhircrackr`, you can use the `httr` package to manually construct and send HTTP requests, and `jsonlite` to convert the JSON response into R data frames. (Note that `fhircrackr` uses `httr` under the hood, but this is hidden with typical use of `fhircrackr`.)

## Setup

```{r setup}
library(tidyverse)

# Used for direct RESTful queries against the FHIR server
library(httr)
library(jsonlite)
```

Set the FHIR server:

```{r}
fhir_server <- "https://api.logicahealth.org/opioids/open"
```

## Example 1: Capability statements

Recall from Exercise 1, we requested the FHIR server's Capability Statement with `fhircrackr::fhir_capability_statement(fhir_server)`

This is how you would accomplish the same request directly with `httr`.
#### Manually querying the FHIR server (without `fhircrackr`)

```{r}
response <- httr::GET(
        url = str_interp("${fhir_server}/metadata"),

        # Note that we request JSON rather than XML from the FHIR server. These are simply different
        # representations of the same underlying data. JSON is easier to work with directly than XML.
        config = list(add_headers( Accept = 'application/fhir+json'))
)

# Convert from raw `httr` response into an R list for easier access
response_list <- fromJSON(httr::content(response, as = "text", encoding = "UTF-8"), flatten = TRUE)
```

Here's what it looks like to pull out the interactions and search parameters from the Capability Statement into R data frames:

```{r}
# Interactions by resource
response_list$rest$resource[[1]] %>%
  mutate(
    interaction = paste(interaction[[1]]$code, collapse = ", ")
  ) %>%
  select(type, interaction)
```
```{r}
# Searches by resource type
response_list$rest$resource[[1]] %>%
  mutate(
    searchParam = paste(searchParam[[1]]$name, collapse = ", ")
  ) %>%
  select(type, searchParam)
```

## Example 2: Simple FHIR search query

In Exercise 1, we request a Bundle of Patient resources from the server with `fhircrackr` like:

```
request <- fhir_url(url = fhir_server, resource = "Patient")
patient_bundle <- fhir_search(request = request)
```

To do this with `httr`:

```{r}
response <- httr::GET(
        url = str_interp("${fhir_server}/Patient"),
        config = list(add_headers( Accept = 'application/json'))
)

# Look at raw JSON
httr::content(response, as = "text", encoding = "UTF-8") %>% prettify()
```

To convert to an R data frame:

```{r}
response_list <- fromJSON(httr::content(response, as = "text", encoding = "UTF-8"), flatten = TRUE)
```

## Example 3: Another FHIR search query

Let's request the MedicationRequest resources for a specific patient from the server:

```{r}
response <- httr::GET(
  url = str_interp("${fhir_server}/MedicationRequest?patient=10098"),
  config = list(add_headers( Accept = 'application/fhir+json'))
)

# Convert from raw `httr` response into an R list for easier access
response_list <- fromJSON(httr::content(response, as = "text", encoding = "UTF-8"), flatten = TRUE)

response_list$entry %>% glimpse
```